---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)


knitr::opts_chunk$set(
  warning    = FALSE,
  fig.align  = "center",
  dpi        = 240,
  fig.width  = 10,
  fig.height = 4.5,
  out.width  = "100%",
  fig.retina = 2,
  dev        = "png",
  cache      = FALSE
)


library(decompositionLE)
library(tidyverse)
```

# decompositionLE

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/decompositionLE)](https://CRAN.R-project.org/package=decompositionLE)
[![Codecov test coverage](https://codecov.io/gh/herts-phei/decompositionLE/graph/badge.svg)](https://app.codecov.io/gh/herts-phei/decompositionLE)
[![R-CMD-check](https://github.com/herts-phei/decompositionLE/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/herts-phei/decompositionLE/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of **decompositionLE** is to provide an easy to use implementation of life expectancy decomposition formulas for age bands, derived from *Ponnapalli, K. (2005)* In addition, there is a decomposition function for disease cause breakdown sourced from *Preston, S.H., Heuveline, P. and Guillot, M. (2001)*, as well as some useful helper plot functions.

## Installation

You can install the development version of decompositionLE from [GitHub](https://github.com/) with:

``` r
devtools::install_github("herts-phei/decompositionLE")
```

## Example

### Age decomposition

The package contains a built-in dataset of life table values for US women born in 1935 and 1995, `us_females`, sourced from *Ponnapalli, K. (2005)*.

```{r}
us_females
```

The dataset demonstrates usage of `decomp_age()`.

```{r}
age_output <- decomp_age(us_females, method = "arriaga3", 
           age_col = "Age", e1 = "e1x", e2 = "e2x", l1 = "l1x", l2 = "l2x")

age_output
```

In `decomp_age()`, there are several methods currently available: `arriaga3`, `chandrasekaran1` or `chandrasekaran2`. Functionally, they produce close to identical results, but the output variables are different.

`age_col` should be a factor column providing ordered age bands with the final age group being an open-ended interval suffixed with '+', e.g. `90+`. `e1` and `l1` correspond to the expectation of life at age group x, and the proportion of persons alive at age group x in decimal form, in the first group of comparison, and likewise for `e2` and `l2` for the second group of comparison.

There is also a helpful plot function for output from `decomp_age()` offering a variety of basic ggplot2 style plots.

```{r, warning=FALSE}
library(patchwork)

plot1 <- plot_age(age_output, method = "arriaga3", plot_type = "segment_dodge")
plot2 <- plot_age(age_output, method = "arriaga3", plot_type = "segment_dodge", line = TRUE)
plot3 <- plot_age(age_output, method = "arriaga3", plot_type = "total", line = TRUE)

plot1
plot2
plot3
```

### Disease decomposition

The package also contains a built-in dataset of age and cause decomposition of difference in Life Expectancies at birth, for India and China, males, 1990, `india_china_males_1990`, sourced from *Murray, C.J.L. and Lopez, A.D. (1996)*.

```{r}
india_china_males_1990
```

The dataset demonstrates usage of `decomp_disease()`.

```{r cols.print=9999, rows.print=9999}
disease_output <- decomp_disease(india_china_males_1990,
  breakdown = "proportion", 
  age_col = "Age",
  diseases = c("CD", "NCD", "Injuries"),
  group_1 = "India", group_1_m = "India_nmx", 
  group_2 = "China", group_2_m = "China_nmx", 
  nDx = "nDx"
)

disease_output
```

The `diseases` argument provides the function with suffixes of relevant diseases found in both groups of interest. `breakdown` is for whether disease breakdowns are `raw` mortality rates or a decimal `proportion` of the total all-cause mortality rate. For example, in the `india_china_males_1990` dataset which has decimal proportion breakdowns:

```{r cols.print=9999, rows.print=9999}
india_china_males_1990[1,]
```

At birth, the components which make up India's all-cause mortality in the dataset:

```{r} 
india_birth_components_equal_1 <- india_china_males_1990[1,c("India_CD", "India_NCD", "India_Injuries")]

all.equal(sum(india_birth_components_equal_1), 1, tolerance = 1e-3)
```

Within tolerance, is equal to 1. Likewise, for China all-cause mortality at birth:

```{r} 
china_birth_components_equal_1 <- india_china_males_1990[1,c("China_CD", "China_NCD", "China_Injuries")]

all.equal(sum(china_birth_components_equal_1), 1, tolerance = 1e-3)
```

However raw disease breakdowns are also accepted if provided, but these disease breakdowns should then equal the value provided in the column for the all-cause mortality rate for the specified group instead, not 1.

Likewise in `decomp_age()`, `age_col` should be a factor column providing ordered age bands with the final age group being an open-ended interval suffixed with '+'.

The `diseases` argument specifies a character vector of the diseases found in both groups, and should be suffixed at the end of the names given in `group_1` and `group_2` in the dataset. In `india_china_males_1990`, these are: `c("CD", "NCD", "Injuries")`.

There is a also a helper plot function for disease decomposition output, `plot_disease()`:

```{r}

plot_disease(x = disease_output, suffixes = c("delta_CD", "delta_NCD", "delta_Injuries"), nDx = "nDx", line = FALSE) 

plot_disease(x = disease_output, suffixes = c("delta_CD", "delta_NCD", "delta_Injuries"), nDx = "nDx", line = TRUE)

```

## References

*Ponnapalli, K. (2005). A comparison of different methods for decomposition of changes in expectation of life at birth and differentials in life expectancy at birth. Demographic Research, 12, pp.141–172. doi: https://doi.org/10.4054/demres.2005.12.7.*

*Preston, S.H., Heuveline, P. and Guillot, M. (2001). Demography : measuring and modeling population processes. Oxford ; Malden, Mass.: Blackwell Publishers.*
